A Constraint Solving Approach to Parikh Images of Regular Languages

### 2024-02-29 example.tex
Fix minor reviewer issues
Closes: #77
Closes: #75
Closes: #76
Closes: #71
Closes: #72
Ensure that figures are refered to, come before ref.
Closes: #78
Closes: #87

\begin{figure}[tb]
  \centering 
  \includegraphics[width=0.6\textwidth]{a_2}
  \caption{ $\mathcal{A}'$ after even more linear algebra, with most transitions expressed
  in terms of $n$. Intuitively, this version captures the fact that $n$ is given
  by distributing the incoming flow of $1$ across the two outgoing transitions
  from $I_{A'}$, the initial state.}
  \label{fig:a_2}
\end{figure}

@@

Note that the per-automata length counting registers have been assigned the same
solver variable $l_A$. Since they have to be equal, either one can be used in
both automata through similar applications of equality elimination rules.

>>>>

\begin{figure}[tb]
  \centering 
  \includegraphics[width=0.6\textwidth]{a_2}
  \caption{ $\mathcal{A}'$ after even more simplification using linear algebra,
   with most transitions expressed
  in terms of $n$. Intuitively, this version captures the fact that $n$ is given
  by distributing the incoming flow of $1$ across the two outgoing transitions
  from $I_{A'}$, the initial state.}
  \label{fig:a_2}
\end{figure}

@@

Note that the per-automata length counting registers have been assigned the same
solver variable~$l_A$. Since they have to be equal, either one can be used in
both automata through similar applications of equality elimination rules.

### 2024-02-29 introduction.tex
Apparently, this is our colour now
Ensure that figures are refered to, come before ref.
Closes: #78
Closes: #87

  The current best published mitigation for this problem relies on the observation
  that any Presburger formula that applies to a product of automata must also hold
  for all of the terms of that product individually. Exploiting this fact, the Parikh
  image of a product of automata is first represented as the conjunction of the image of
  the individual automata of the product, falling back to computing the entire product
  for satisfiable cases~\cite{approximate-parikh}.

@@

Our primary influence for this work is the operation of (finite-domain) 
constraint programming (CP)%
~solvers~\cite{cp}. CP~solvers typically employ a combination of lazy enforcement
of constraints through \emph{propagation},
lazily pruning values from decision variable domains that do not satisfy 
the constraints during solving, and \emph{branching}, splitting proofs
by cases in a way that will encourage maximum propagation.

For our problem, the constraints are automata connectivity and linear
equations (Presburger formulas).
We use propagation to lazily enforce connectivity of automata,
removing transitions as they become unreachable during solving or are
eliminated by constraints in the Presburger formula. We
also branch on the presence or absence of transitions to
encourage propagation of the connectivity constraint. This allows
us to whittle down the automata
(including intermittent products) before moving on with potentially
 explosive product construction. The underlying observation
is that though the problem domain and the domain of our integer variables
is infinite, the number of simple paths
through the automata (and their product) is finite. In summary, the key ideas are:

{
    \centering
    \begin{tcolorbox}[colback=gray!5!white,colframe=gray!75!black,%
        title=Key ideas of \Calculus{},%
        width=0.8\linewidth]

@@

We extend the \Ostrich{} string solver's backend portfolio with \Calculus{} via
\Catra{} as the Parikh image intersection solver of the \OstrichPlus{}~algorithm,
obtaining improvements in both \textsc{Sat}  and \textsc{Unsat} performance in
\Ostrich{}' results from SMT-COMP 2023, crucially allowing \Ostrich{} to win
the unsatisfiable category of the track of quantifier-free, linear constraints
over string logic (QF\_SLIA) and increasing the number of \textsc{Sat}~results
by \numprint{3}\%.

>>>>

  The current best published mitigation for this problem relies on the observation
  that any Presburger formula that applies to a product of automata must also hold
  for all of the terms of that product individually. Exploiting this fact, the Parikh
  image of a product of automata is first represented as the conjunction of the image of
  the individual automata of the product, falling back to computing the entire product
  for satisfiable cases or cases when unsatisfiability is only discoverable after
  full- or partial product construction~\cite{approximate-parikh}.

@@

Our primary influence for this work is the operation of (finite-domain) 
constraint programming (CP)%
~solvers~\cite{cp}. CP~solvers typically employ a combination of lazy enforcement
of constraints through \emph{propagation},
pruning values from decision variable domains that do not satisfy 
the constraints during solving, and \emph{branching}, splitting proofs
by cases in a way that will encourage maximum propagation.

For our problem, the constraints are automata connectivity and linear
equations (Presburger formulas).
We use propagation to lazily enforce connectivity of automata,
removing transitions as they become unreachable during solving or are
eliminated by constraints in the Presburger formula. We
also branch on the presence or absence of transitions to
encourage propagation of the connectivity constraint. This allows
us to whittle down the automata
(including intermittent products) before moving on with potentially
 explosive product construction. The underlying observation
is that though the domain of our integer variables
is infinite, the number of simple paths
through the automata (and their product) is finite. In summary, the key ideas are:

{
    \centering
    \begin{tcolorbox}[colback=ourcolour!5!white,colframe=ourcolour!75!black,%
        title=Key ideas of \Calculus{},%
        width=0.8\linewidth]

@@

We extend the \Ostrich{} string solver's backend portfolio with \Calculus{} via
\Catra{} as the Parikh image intersection solver of the \OstrichPlus{}~algorithm,
obtaining improvements in both \textsc{Sat}  and \textsc{Unsat} performance in
\Ostrich{}' results from SMT-COMP 2023, crucially allowing \Ostrich{} to win
the unsatisfiable category of the track of quantifier-free, linear constraints
over string logic (QF\_SLIA) and increasing the number of \textsc{Sat}~results
by \numprint{3}\% on the same track.

### 2024-02-29 multiple.tex
Ensure that figures are refered to, come before ref.
Closes: #78
Closes: #87

We then continue by repeating the steps of \cref{sec:single:example} shown in
\cref{fig:derivation:single}, since the steps to propagate
\cref{ex:string-constraints} across the automaton $\mathcal{A}'$ still apply for the
product. Note that this fixes the values of all transitions but one of $\mathcal{A}'$ as
constants, including the free counter variables $i, n$, whose new constant
values we propagate everywhere.

In both cases, we have comparatively simple automata left, so we apply
\Materialise{} to obtain their product, which for both branches is empty. An
empty product will have no outgoing transitions from its initial state, and so
will produce a flow equation $1 = 0$ since the right-hand balancing side will
always be the empty sum. We can then close the proof and know that the problem
is unsatisfiable.

@@

>>>>

@@

We then continue by repeating the steps of~\cref{sec:single:example}, shown 
in~\cref{fig:derivation:single} since the steps to propagate
\cref{ex:string-constraints} across the automaton $\mathcal{A}'$ still apply for the
product. Note that this fixes the values of all transitions but one of $\mathcal{A}'$ as
constants, including the free counter variables $i, n$, whose new constant
values we propagate everywhere.

In both cases, we have comparatively simple automata left, so we apply
\Materialise{} to obtain their product, which for both branches is empty. An
empty product will have no outgoing transitions from its initial state, and so
will produce a flow equation $1 = 0$ since the right-hand balancing side will
always be the empty sum. We can then close the proof and know that the problem
is unsatisfiable. A full derivation tree for the example can be found 
in~\cref{fig:derivation:multi}.
